# pre-class


Make sure you commit this often with meaningful messages. 



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```








Standardizing a variable means subtracting the mean, and then dividing by the standard deviation. Let’s use a loop to standardize the numeric columns in the [Western Collaborative Group Study](https://clinicaltrials.gov/ct2/show/NCT00005174). This study began in 1960 with 3154 men ages 39-59, who were employed in one of 11 California based companies. They were followed until 1969 during this time, 257 of these men developed coronary heart disease (CHD). You can read this data in with the code below. You can access this dataset with the following code:

```{R}
suppressMessages(library(foreign))
wcgs <- read.dta("https://drive.google.com/uc?export=download&id=0B8CsRLdwqzbzYWxfN3ExQllBQkU")
install.packages("dplyr")
library("dplyr")

```

The data has the following variables:



WCGS has the following variables:

-----------------------------------------------------------
Name    Description
------- -------------------------------------------
id      Subject identification number

age     Age in years

height  Height in inches

weight  Weight in lbs.

sbp     Systolic blood pressure in mm 

dbp     Diastolic blood pressure in mm Hg

chol    Fasting serum cholesterol in mm 

behpat  Behavior

  1       A1

  2       A2

  3       B3

  4       B4

ncigs   Cigarettes per day

dibpat  Behavior

1       type A

2       type B

chd69   Coronary heart disease

1       Yes

0       no

typechd Type of CHD

1       myocardial infarction or death

2       silent myocardial infarction

3       angina perctoris

time169 Time of CHD event or end of follow-up

arcus   Arcus senilis

0       absent

1       present

bmi     Body Mass Index
-----------------------------------------------------------




### Question 1: Standardize Function

A. Create a function called standardize.me() that takes a numeric vector as an argument, and returns the standardized version of the vector. 

```{r}
# defining a function stardrize me that takes on argument x. Set na.rm=TRUE to get remove NAs
standarize.me<- function(x){
  (x-mean(x,na.rm=TRUE))/sd(x,na.rm=TRUE)
}

```


B. Assign all the numeric columns of the original WCGS dataset to a new dataset called WCGS.new.
```{r}
# select all columns that are numeric in wcgs, and save it as WCGS.new
WCGS.new<-wcgs%>%
  select(age:chol,ncigs,bmi,time169,agecat) 
WCGS.new
```

C. Using a loop and your new function, standardize all the variables WCGS.new dataset.
```{r}
#loop through columns of WCGS.new
for (i in 1:ncol(WCGS.new)) {
  # standarized each value in column i 
  WCGS.new[,i]<-standarize.me(WCGS.new[,i])
  }
WCGS.new
```
D. What should the mean and standard deviation of all your new standardized variables be? Test your prediction by running a loop

You would expect the mean to be zero and the standard deviation to be 1 because the variabels are standarized and centered. 

```{r}
# iterate through the columns of WCGS.new
for (i in 1:ncol(WCGS.new)){
# create new variable mean_wcgs, which takes the mean of each column, removing NAs
  mean_wcgs<-round(mean(WCGS.new[,i],na.rm = TRUE))
# pring the mean mean_wcgs for each column i 
  print(paste("the mean is", mean_wcgs))
# Do the same thing for sds
  sd_wcgs<-sd(WCGS.new[,i],na.rm = TRUE)
  print(paste("the sd is", sd_wcgs))
}
WCGS.new
```


### Question 2: Looping to Calculate

A. Using a loop, calculate the mean weight of the subjects separated by the type of CHD they have.
```{r}

# make variable that saves all unique values from typchd69 as a character (no CHD, Mi or SD etc.)
Groups = as.character(unique(wcgs$typchd69))

# Create empty data frame with variables Groups and Mean_weight, where Group the unique types of CHD 
# 
Means = data.frame(
  Group = Groups,
  Mean_Weight = rep(NA, length(Groups)) 
)

# Iterate through the groups column
for (j in  1:length(Groups)) {
    # Create empty vector called Group_mean
  Grp_mean = vector()
    # Iterate through the rows of wcgs
 for(i in 1:nrow(wcgs)){
   # if the chd in row i is equal to the group in row j you are looking for (no CHD, MI or SD etc)
   if(wcgs$typchd69[i] == Groups[j]){
    # then Add the corresponding weights to the group mean vector 
     Grp_mean = c(Grp_mean , wcgs$weight[i])}}
    # save mean of the the group mean vector in corresponding position in means
 Means[j,2] = mean(Grp_mean)}

Means

```

B. Now do the same thing, but now don’t use a loop

```{r}

# Group by typchd69 and then summarise by getting the mean of weight 
names(wcgs)
wcgs%>%
  group_by(typchd69)%>%
  summarise(mean(weight))

```




